\replaced[id=siyu]{Separating random and deterministic sources of computational noises in explore-exploit decisions}{The nature of decision noise in random exploration}

### 2022-12-21 manuscript/2022 draft - Plos Comp Bio/manuscript_DeterministicRandomNoise.tex
siyu final draft

	\newpage
	\begin{abstract}
Human decision making is inherently variable. While this variability is often seen as a sign of suboptimality behavior, recent work suggests that variability can actually be adaptive. An example arises when we must choose between exploring unknown options or exploiting options we know well. A little randomness in these `explore-exploit' decisions is remarkably effective as it encourages us to explore options we might otherwise ignore. Recent work suggests that people may actually use such `random exploration' in practice, increasing their behavioral variability when it is more valuable to explore. 
%Other work, however, suggests that variability in human behavior is not always random, and is instead driven by deterministic factors (such as repeating past actions) 
A key question is whether the variability in random exploration is actually random. That is, is random exploration driven by stochastic processes in the brain or by some unobserved deterministic process that we have failed to account for when measuring behavioral variability? By designing an explore-exploit task in which, unbeknownst to them, participants are presented with the exact same choice twice, we provide a partial answer to this question. \added[id=siyu]{By a novel bayesian analysis, we are able to quantify the relative magnitudes of stimulus-driven deterministic noise and non-stimulus-driven random noise in driving random exploration.} In particular, we find evidence that at least 20$\%$ of the variability in random exploration can \added[id=siyu]{indeed} be accounted for by deterministic processing of the stimulus. This still leaves open the possibility that much of random exploration is truly `random', but narrows the window of opportunity for \added[id=siyu]{true} stochastic processes to explain this behavior. \added[id=siyu]{Our work sheds light on how random exploration may be implemented by reducing the filtering of task-irrelevant stimuli in the decision process.}

@@

	\replaced[id=siyu]{One limitation, however, is that it is difficult to tell whether what we have called 'decision noise' in previous research is truly a random noise.}{One limitation of this previous work, is that it is difficult to tell whether what is measured as decision noise is truly random.} That is, whether behavioral variability is due to intrinsic stochastic processes in the brain or whether it is due to deterministic processes that we \replaced[id=siyu]{failed to observe}{simply have not observed}. \added[id=siyu]{Decision noise as defined in previous researches are more or less a quantification of what's not predictable by the model. A missing deterministic component from the model could give rise to variability in behavior that might be mistakenly taken as a random noise.} For example, in the restaurant example, my usual preference for one restaurant or another may be overruled if I see an ex romantic partner going into one of them. Avoiding an ex is a deterministic process, but if we fail to take ex's presence into account as scientists modeling the decision, then over a series of such decisions where the ex is present or not, we would mistakenly attribute the ensuing `variability' in choice to randomness.

@@

	In this paper, we investigate the extent to which the apparent randomness in random exploration can be explained by such \added[id=siyu]{unobserved} deterministic processing of the stimulus \added[id=siyu]{(will refer to as deterministic noise)}. \replaced[id=siyu]{To distinguish between stimulus-driven `deterministic noise' and non-stimulus-driven `random noise'}{In particular}, we modify the Horizon Task \citep{wilson2014} to have people face the exact same explore-exploit choice twice. If the decision is a purely deterministic function of the stimulus \added[id=siyu]{(i.e., decision noise is purely deterministic noise)}, then people's choices should be identical both times. That is, their choices should be consistent, since the stimulus is the same both times. Conversely, the more their decision is driven by \replaced[id=siyu]{random noise}{other processes, including both stochastic and unobserved deterministic processes}, the less consistent their behavior should be. By analyzing behavior on this task in both a model-free and model-based manner, we show that at least some of the `randomness' in random exploration must come from deterministic processing of the stimulus. \replaced[id=siyu]{Our work}{This does not prove that random exploration is entirely deterministic or stochastic, but} provides a lower bound on how much deterministic processes contribute to random exploration\added[id=siyu]{, and an upper bound on how much true random noises contribute to random exploration}. \deleted[id=siyu]{This in turn sheds light on how random exploration may be implemented by amplifying the effects of task-irrelevant stimuli on choice.}

@@

	Relative to the original Horizon Task, the key modification here is to give people `repeated games,' in which they see exact same set of example plays twice in two separate games (separated by several minutes in time so as to avoid detection). By repeating the instructed plays for each game twice, we can set up a situation where (unbeknownst to the participants) they are faced with the exact same explore-exploit choice, with the exact same stimuli twice. Thus, if their behavior is a deterministic function of the stimuli, they will behave identically on the two games, that is their \replaced[id=siyu]{choice}{behavior} on the two versions of each game will be consistent. Conversely, if their behavior is not driven by a deterministic function of the stimulus, then their choices on the repeated games \replaced[id=siyu]{could}{should} be inconsistent. \replaced[id=siyu]{The extent to which participants' choices are consistent on the repeated versions of the games allow}{On average participants played 67.31 repeated games each allowing} us to quantify the extent to which their behavior was \replaced[id=siyu]{driven by a deterministic noise process vs a random noise process}{a deterministic function of the stimulus or not}.

@@

	\subsubsection*{\replaced[id=siyu]{Model validation}{Parameter recovery}\label{ch:appendix:bayesrecovery}}
	To be sure that our fit parameter values were meaningful \added[id=siyu]{and to understand the limits of our model, we evaluated our model in several ways. Firstly, we checked if our fitted deterministic noise could indeed capture unobserved deterministic process that was not accounted for by the decision model. In particular, we fit a reduced version of our model that only considers reward and ignores the influence of uncertainty condition on explore-exploit decisions.  $\Delta Q, \Delta R, n_{det}, n_{ran}$ below represent the same variables as in the full model.
	$$\Delta Q= \Delta R+n_{det}+n_{ran}$$
	If deterministic noise in our model can indeed capture unobserved deterministic processes that's missed by the model, then we would expect to see a higher level of fitted deterministic noise in the reduced model compared to in the full model, whereas the level of random noise should remain unchanged. By comparing the	fitted posterior distributions over the group-level means of the deterministic and random noise parameters $\sigma_{det}$ and $\sigma_{ran}$, as expected, we observed an increase in deterministic noise and no change in random noise between the reduced and the full model (Figure \ref{fig:reducedmodel}). 
	}
	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=0.7\textwidth]{figures/RDBayes_reduced_model.jpg}
			\caption{Deterministic noise can recover known deterministic processes that's intentionally omitted by the model. In the reduced model where the deterministic effect of uncertainty condition is omitted from the model, deterministic noise is higher compared to the full model that accounts for the effect of uncertainty. Random noise remains unchanged between the two models.}
			\label{fig:reducedmodel}
		\end{center}
	\end{figure}

	\added[id=siyu]{Secondly, we evaluated our hierarchical Bayesian analysis procedure using the `frequentist coverage analysis'. In the coverage analysis, we simulated choices with the fitted parameters from the Hierarchical Bayesian analysis, and the re-fit the simulated choices to see whether we can recover the parameters (Figure \ref{fig:coverage2}, Supplementary Figure S4). The simulation and re-fitting was repeated for 200 times. Then we counted out of the 200 repetitions, whether the true parameter that we simulate the choices from lies in the fitted 95\% confidence interval for more than 95\% of the times (this ratio will be referred to as the coverage rate). For random noise, the coverage rate is 100\% for both horizon 1, horizon 6 and noise differences between horizons. For deterministic noise, the coverage rate is 66\% for horizon 1, 69\% for horizon 6, and 97\% for noise differences between horizons. By comparing the posterior distributions of parameters that were used to generate simulations and the posterior distribution of re-fitted parameters to the simulated data, it is clear that our model systematically underestimates deterministic noise. For random noise, our model fitting procedure yields a faithful recovery. However, because our random noise is modeled as non-stimulus-driven noise which includes both truly stochastic neural noise and possible deterministic noises irrelevant to the stimuli, our random noise estimate provides an upper bound of `random noise' induced by intrinsic stochastic processes in the brain. Because the underestimation of deterministic noise is partially canceled out when taking the difference, our model fitting procedure passed the coverage test in estimating the horizon differences in both random and deterministic noise.}

@@

	\added[id=siyu]{Next, we tested the ability of our model fitting procedure to recovery parameters from simulated data at the subject level (Supplementary Figure S5 and S6). The correlations between the true vs fitted parameters are significant across participants for all parameters (p $<$ 0.001). Despite the strong correlation between simulated and fit values of deterministic noise (R $>$ 0.8), we again observed a systematic underestimation of $\sigma_{det}$ (Supplementary Figure S5 and S6).}

@@

	\added[id=siyu]{Lastly, in addition to testing how our model performs in parameter ranges around the actual fitted parameters, we tested the limitations of our models in arbitrary combinations of random vs deterministic noises. All combinations of random and deterministic noises with $0 \le \sigma_{det} \le 10$ and $0 \le \sigma_{ran} \le 10$ are simulated. In a special case, we evaluated how our model performs when there is only random noise or only deterministic noise (Figure \ref{fig:puredetran}). In the simulation with fully deterministic noise and 0 random noise, our model successfully recovered both random and deterministic noise (Figure \ref{fig:puredetran} C, D), however in the simulation with full random noise and 0 deterministic noise, although our model successfully recovered random noise, some small proportion of deterministic noise was falsely detected when they should instead be 0 (Figure \ref{fig:puredetran} A, B). However, this phenomenon only exists when the true deterministic noise is 0, once the true deterministic noise is greater than 1, we don't observe this inflation of deterministic noise anymore (Supplementary Figures S7). Overall, our model did a fairly good job in recovering all combinations of random and deterministic noises (Supplementary Figures S7).}

	%This is because it requires more trials to recover larger values of $\sigma_{det}$ and $\sigma_{ran}$, so with the same number of choices it is harder to recover overall larger noise variances in horizon 6. In addition we see better recovery for random noise than deterministic noise. This is likely because we effectively have half as many trials for deterministic noise. In particular, while we generate two samples of random noise for each repeated game pair, we only generate one sample of deterministic noise, which by definition is the same in both of the repeated games.

@@

	\added[id=siyu]{Overall, we are able to detect both deterministic and random noises using our model to a satisfactory extent. We see better parameter recovery for random noise than deterministic noise. This is likely because we effectively have half as many trials for deterministic noise. In particular, while we generate two samples of random noise for each repeated game pair, we only generate one sample of deterministic noise, which by definition is the same in both of the repeated games. Our model provides a lower bound for deterministic noise and an upper bound for random noise.
	}
	\subsubsection*{Model-based results}

	Posterior distributions over the group-level means of the deterministic and random noise standard deviation $\sigma_{det}$ and $\sigma_{ran}$ are shown in Figure \ref{fig:mb1} and Supplementary Figure S8. Consistent with our model-free results, we see that both random and deterministic noise are non-zero. \replaced[id=siyu]{Numerically,}{and that} random noise is about 2-3 times larger than the deterministic noise. \added[id=siyu]{By computing the 5\% quantile of the posterior distribution of $\sigma_{det}/(\sigma_{det}+\sigma_{ran})$, our data suggests that at least 20.05\% of the variability in random exploration can be accounted for by deterministic noise.} In addition, we find that both random and deterministic noise increase with horizon. This increase was larger for random noise (M = 7.13, 100\% of samples showed an increase in random noise with horizon) than deterministic noise (M = 2.59, 98.64\% of samples showed an increase in deterministic noise with horizon). But intriguingly the relative increase in both types of noise was \replaced[id=siyu]{similar}{the same} (Figure \ref{fig:ratio}). That is when we compute the relative increase in deterministic noise with horizon, $\sigma^{det}_{horizon6}/\sigma^{det}_{horizon1}$, it is \replaced[id=siyu]{very similar}{almost identical} to the relative increase in random noise with horizon $\sigma^{ran}_{horizon6}/\sigma^{ran}_{horizon1}$. 
	%BOB SAYS: CHECK NOTATION HERE AND INSERT FIGURE SHOWING RATIO

@@

	% BOB SAYS: CAN YOU TRY TO REFRAME THIS AROUND THE DETERMINISTIC STORY
	\added[id=siyu]{Randomness in exploratory behavior can be driven by both the intrinsic variability generated in the brain (show up in behavior as random noise) and by deterministic neural processes induced by the stimuli (show up in behavior as deterministic noise).} In this paper, we investigated whether random exploration is really random or whether it is driven deterministically by aspects of the stimulus we have previously ignored when measuring `decision noise'.  Using a version of the Horizon Task with repeated games, we found evidence that at least some of the noise in random exploration could be explained by such `deterministic noise.' In particular, we found that deterministic noise accounted for at least 20.05\% of the overall variability in people's behavior.

@@

	So where does this leave randomness in random exploration? \replaced[id=siyu]{In our model, random noise is modeled as non-stimulus-driven variability, which may include truly random noise (intrinsic stochasticity in decision making) or stimulus irrelevant deterministic processes that we could not control in the lab. As a result,}{Well, } the remaining 79.95 \% of the decision noise could be random or it could be deterministic \replaced[id=siyu]{processes that is not bound to the stimulus and that we can not control}{because we can't control everything}. In particular, while we controlled many aspects of the stimulus across repeated games (e.g. the outcomes and the order of the forced trials), we could not perfectly control {\it all} stimuli the participant received, which would vary, for example, based on exactly what they were looking at or whether they were fidgeting or scratching their nose \citep{Musall2019}. Thus conceptually, our estimate of random noise is an upper bound as these `missing' sources of deterministic noise would be interpreted as random noise in our model. Conversely, our estimate of deterministic noise is a lower bound. \added[id=siyu]{Future work is needed to identify these additional sources of deterministic noise that are not controlled in our work, for example by tracking people's pose, body movements and eye movements during the experiment.}

	In addition to the conceptual limitation in measuring deterministic noise, parameter recovery simulations suggest that our estimation method also systematically underestimates deterministic noise (see Figure \ref{fig:coverage2}, Supplementary Figure S6). From both a conceptual and methodological perspective, our model provides a lower bound of deterministic noise and an upper bound of random noise. Although there is still a considerable window for truly stochastic processes in the brain to be driving random exploration (up to 79.95\%), our results suggest that at least some of the apparent randomness in random exploration is not random at all.

	\added[id=siyu]{This is in line with works of \citep{Bair1996} in which MT neurons were shown to have a reproducible temporal modulation in response to a fixed random motion stimuli. In other words, irrelevant features in the stimuli are represented in a reliable way in the brain that could drive downstream choices in a predictable way. A more recent work showed that neural variability in the visual areas reflect statistics in the stimuli images \citep{Festa2021}. Of particular interest was the fact that deterministic noise in our study increased with horizon, which is a hallmark of an exploratory process and suggests that the modulation of deterministic processes may underlie random exploration. }

	\added[id=siyu]{Other research are in line with random exploration \citep{drugowitsch16}. In particular these authors show that randomness in behavior arises from imperfections in mental inference, that happen inside the brain, rather than in peripheral processes such as sensory processing and response selection. Although there is a distinction between suboptimal inference and random noise \citep{Pouget12}, they both show up as random noise in our definition. This suggests that a large proportion of variability in behavior may arise from computational errors in computing the correct strategy. In the context of the Horizon Task, such computational errors would likely be larger in the long horizon condition as the correct course of action in these cases is much harder to compute.	
	}
	%this is consistent with findings of \cite{drugowitsch16}. In particular these authors show that randomness in behavior arises from imperfections in mental inference, that happen inside the brain, rather than in peripheral processes such as sensory processing and response selection. This suggests that a large proportion of noise in behavior is generated randomly and that this may arise from computational errors in computing the correct strategy. % In the context of the Horizon Task, such computational errors would likely be larger in the long horizon condition as the correct course of action in these cases is much harder to compute.

@@

	%Taken at face value, the horizon-dependent increase in random noise is consistent with the idea that random exploration is driven by intrinsic variability in the brain. 
	\added[id=siyu]{
	Regardless of whether the remaining 79.95 \% is deterministic or random, the fact that the horizon change in the two noises are proportional to each other (Figure \ref{fig:ratio}) suggests a possible mechanism for random exploration. Specifically, a reduction in the strength with which reward drives the choice. In our decision model, choice is determined by the sign of the difference in utility $\Delta Q$ between the two options, thus, the absolute value of different terms in the model do not matter, $\Delta Q$ is only affected by the relative magnitude of reward, information, bias and noise. Mathematically, our model is equivalent to
	}
	\begin{equation*}
		\begin{split}
			\Delta Q' &= \beta \Delta Q \\
			&= \beta \Delta R + \beta A \Delta I + \beta b + \beta n_{det} + \beta n_{ran}\\
			&= \beta \Delta R + A' \Delta I + b' + n_{det}' + n_{ran}'
		\end{split}
	\end{equation*}
	\added[id=siyu]{
	In this equivalent form, $\beta$ determines the weight given to the reward in the decision. From a model fitting perspective, these two forms are equivalent. A decrease in $\beta$ here would be equivalent to an increase in variance of both deterministic and random noise. From a psychological perspective, however, they are quite different. An increase in both random and deterministic noise at a similar ratio, as observed in our experiment when horizon increases, suggests that either $\beta$ is reduced or $n_{det}'$ and $n_{ran}'$ are increased simultaneously at the same ratio. Neurally, reducing the $\beta$ corresponds to a reduction in reward sensitivity and reward coding, whereas simultaneously increasing $n_{det}'$ and $n_{ran}'$ corresponds to a reduction in some common noise filtering mechanism that reduces both random and deterministic noises. }

@@

	%SIYU says the song bird literature supports that behavioral variability is reflected in neural variability. But it can't tell whether the neural variability arises from deterministic processes of external stimuli or random noise. I didn't read/remember the songbird paper, this is only based on the text below.
			%"The idea that random exploration is driven by truly stochastic processes in the brain is in line with work in the bird song literature in which variability during song learning has been tied to neural variability arising from specific areas of the brain \citep{songbird1, songbird2}"

>>>>

	\newpage
	\begin{abstract}
Human decision making is inherently variable. While this variability is often seen as a sign of suboptimality behavior, recent work suggests that variability can actually be adaptive. An example arises when we must choose between exploring unknown options or exploiting options we know well. A little randomness in these `explore-exploit' decisions is remarkably effective as it encourages us to explore options we might otherwise ignore. Recent work suggests that people may actually use such `random exploration' in practice, increasing their behavioral variability when it is more valuable to explore. 
%Other work, however, suggests that variability in human behavior is not always random, and is instead driven by deterministic factors (such as repeating past actions) 
A key question is whether the variability in random exploration is actually random. That is, is random exploration driven by stochastic processes in the brain or by some unobserved deterministic process that we have failed to account for when measuring behavioral variability? By designing an explore-exploit task in which, unbeknownst to them, participants are presented with the exact same choice twice, we provide a partial answer to this question. \added[id=siyu]{We built a novel computational model which could detect and separate previously unobserved deterministic processes from random noise, without having to make a priori assumptions about the nature of the deterministic process. } In particular, we find evidence that around 15$\%$ of the variability in random exploration can be accounted for by deterministic processing of the stimulus. This still leaves open the possibility that much of random exploration is truly `random', but narrows the window of opportunity for \added[id=siyu]{true} stochastic processes to explain this behavior. \added[id=siyu]{Our work takes us one step closer to understanding of the nature of decision noise in random exploration.}

@@

	\replaced[id=siyu]{One limitation, however, is that it is difficult to tell whether what we have called 'decision noise' in previous research is truly a random noise.}{One limitation of this previous work, is that it is difficult to tell whether what is measured as decision noise is truly random.} That is, whether behavioral variability is due to intrinsic stochastic processes in the brain or whether it is due to deterministic processes that we \replaced[id=siyu]{failed to observe}{simply have not observed}. \added[id=siyu]{Decision noise as defined in previous researches are more or less a quantification of what's not predictable by the model. A missing deterministic component from the model could give rise to variability in behavior that might appear to be a random noise.} For example, in the restaurant example, my usual preference for one restaurant or another may be overruled if I see an ex romantic partner going into one of them. Avoiding an ex is a deterministic process, but if we fail to take ex's presence into account as scientists modeling the decision, then over a series of such decisions where the ex is present or not, we would mistakenly attribute the ensuing `variability' in choice to randomness.

@@

	In this paper, we investigate the extent to which the apparent randomness in random exploration can be explained by such \added[id=siyu]{unobserved} deterministic processing of the stimulus \added[id=siyu]{(will refer to as deterministic noise)}. \replaced[id=siyu]{To distinguish between stimulus-driven `deterministic noise' and non-stimulus-driven `random noise'}{In particular}, we modify the Horizon Task \citep{wilson2014} to have people face the exact same explore-exploit choice twice. If the decision is a purely deterministic function of the stimulus \added[id=siyu]{(i.e., decision noise is purely deterministic noise)}, then people's choices should be identical both times. That is, their choices should be consistent, since the stimulus is the same both times. Conversely, the more their decision is driven by \replaced[id=siyu]{random noise}{other processes, including both stochastic and unobserved deterministic processes}, the less consistent their behavior should be. \added[id=siyu]{We built a computational model to quantify the relative magnitudes of stimulus-driven deterministic noise and non-stimulus-driven random noise in driving random exploration. Our model could detect the existence of previously unobserved deterministic processes without making a priori assumptions about the deterministic process that we want to observe.} By analyzing behavior on this task in both a model-free and model-based manner, we show that at least some of the `randomness' in random exploration must come from deterministic processing of the stimulus. \replaced[id=siyu]{Our work}{This does not prove that random exploration is entirely deterministic or stochastic, but} provides a lower bound on how much deterministic processes contribute to random exploration\added[id=siyu]{, and an upper bound on how much true random noises contribute to random exploration}. \deleted[id=siyu]{This in turn sheds light on how random exploration may be implemented by amplifying the effects of task-irrelevant stimuli on choice.}

@@

	Relative to the original Horizon Task, the key modification here is to give people `repeated games,' in which they see exact same set of example plays twice in two separate games (separated by several minutes in time so as to avoid detection). By repeating the instructed plays for each game twice, we can set up a situation where (unbeknownst to the participants) they are faced with the exact same explore-exploit choice, with the exact same stimuli twice. Thus, if their behavior is a deterministic function of the stimuli, they will behave identically on the two games, that is their \replaced[id=siyu]{choices}{behavior} on the two versions of each game will be consistent. Conversely, if their behavior is not driven by a deterministic function of the stimulus, then their choices on the repeated games \replaced[id=siyu]{could}{should} be inconsistent. \replaced[id=siyu]{The extent to which participants' choices are consistent on the repeated versions of the games allow}{On average participants played 67.31 repeated games each allowing} us to quantify the extent to which \added[id=siyu]{the variability in} their behavior was \replaced[id=siyu]{driven by a deterministic process vs a random noise process}{a deterministic function of the stimulus or not}.

@@

	\subsubsection*{\replaced[id=siyu]{Model validation}{Parameter recovery}\label{ch:appendix:bayesrecovery}}
	To be sure that our fit parameter values were meaningful \added[id=siyu]{and to understand the limits of our model, we evaluated our model in several ways. Firstly, we checked if our fitted deterministic noise could indeed capture unobserved deterministic process that was not accounted for by the decision model. We test this by leaving out one known deterministic process from the decision model, and ask if our method could recover that known deterministic process as deterministic noise. In particular, we fit a reduced version of our model that only considers reward and ignores the influence of uncertainty condition on explore-exploit decisions.  
	$$\Delta Q= \Delta R+n_{det}+n_{ran}$$
	Here, $\Delta Q, \Delta R, n_{det}, n_{ran}$ represent the same variables as in the full model.	If deterministic noise in our model can indeed capture unobserved deterministic processes that's missed by the model, then we would expect to see a higher level of fitted deterministic noise in the reduced model compared to in the full model, whereas the level of random noise should remain unchanged. By comparing the	fitted posterior distributions over the group-level means of the deterministic and random noise parameters $\sigma_{det}$ and $\sigma_{ran}$, as expected, we observed an increase in deterministic noise and no change in random noise between the reduced and the full model (Figure \ref{fig:reducedmodel}). This suggests that our model is capable of detecting missing deterministic processes.
	}
	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=0.7\textwidth]{figures/RDBayes_reduced_model.jpg}
			\caption{Deterministic noise can recover known deterministic processes that's intentionally omitted by the model. In the reduced model where the deterministic effect of uncertainty condition is omitted from the model, deterministic noise is higher compared to the full model that accounts for the effect of uncertainty. Random noise remains unchanged between the two models.}
			\label{fig:reducedmodel}
		\end{center}
	\end{figure}

	\added[id=siyu]{Secondly, we evaluated our hierarchical Bayesian analysis procedure using the `frequentist coverage analysis'. In the coverage analysis, we simulated choices with the fitted parameters from the Hierarchical Bayesian analysis, and then re-fit the simulated choices to see whether we can recover the parameters (Figure \ref{fig:coverage2}, Supplementary Figure S4). The simulation and re-fitting was repeated for 200 times. Then we counted out of the 200 repetitions how many times the true parameter that we simulate the choices from lies in the fitted 95\% confidence interval. If our model fitting is reliable, then the fitted confidence interval should cover the true parameter for more than 95\% of the simulations (this ratio will be referred to as the coverage rate). For random noise, the coverage rate is 100\% for both horizon 1, horizon 6, and the horizon difference. For deterministic noise, the coverage rate is 66\% for horizon 1 and 69\% for horizon 6. By comparing the posterior distributions of parameters that were used to generate simulations and the posterior distribution of recovered parameters, it is clear that our model systematically underestimates deterministic noise (Figure \ref{fig:coverage2}). Despite the underestimation of deterministic noise in both horizons, we could still reliably detect the horizon changes of deterministic noise (coverage rate is 97\%). This is because the underestimation of deterministic noise is partially canceled out when the difference is taken between horizons. For random noise, our model fitting procedure yields a faithful recovery. However, there is a conceptual limitation. Because random noise is modeled as non-stimulus-driven noise, it can include both true stochastic neural noise and possible deterministic noises which do not depend on the stimuli. Because of this, our random noise estimate provides an upper bound of true `random noise' induced by intrinsic stochastic processes in the brain.}

@@

	\added[id=siyu]{Next, we tested the ability of our model fitting procedure to recovery parameters from simulated data at the subject level (Supplementary Figure S5 and S6). The correlations between the true vs fitted parameters are significant across participants for all parameters (p $<$ 0.001). The strength of correlation between simulated and fit values are strong for both deterministic noise (R $>$ 0.8) and random noise (R $>$ 0.9). Despite the strong inter-subject correlations, we again observed a systematic underestimation of $\sigma_{det}$ (Supplementary Figure S5 and S6).}

@@

	\added[id=siyu]{Lastly, in addition to testing how our model performs in parameter ranges around the actual fitted parameters, we tested the limitations of our models in arbitrary combinations of random vs deterministic noises. All combinations of random and deterministic noises with $0 \le \sigma_{det} \le 10$ and $0 \le \sigma_{ran} \le 10$ were tested. In a special case, we evaluated how our model performs when there is only random noise or only deterministic noise (Figure \ref{fig:puredetran}). In the simulation with fully deterministic noise and 0 random noise, our model successfully recovered both random and deterministic noise (Figure \ref{fig:puredetran} C, D), however in the simulation with fully random noise and 0 deterministic noise, although our model successfully recovered random noise, some small proportion of deterministic noise was falsely detected when they should instead be 0 (Figure \ref{fig:puredetran} A, B). However, this phenomenon only exists when the true deterministic noise is 0, once the true deterministic noise is greater than 1, we don't observe this inflation of deterministic noise anymore (Supplementary Figures S7). Apart from this, our model did a fairly good job in recovering all combinations of random and deterministic noises (Supplementary Figures S7).}

	%Overall, we see better recovery for random noise than deterministic noise. This is likely because we effectively have half as many trials for deterministic noise. In particular, while we generate two samples of random noise for each repeated game pair, we only generate one sample of deterministic noise, which by definition is the same in both of the repeated games.

	%This is because it requires more trials to recover larger values of $\sigma_{det}$ and $\sigma_{ran}$, so with the same number of choices it is harder to recover overall larger noise variances in horizon 6.

@@

	\added[id=siyu]{Overall, we are able to detect both deterministic and random noises using our model to a satisfactory extent. Our model provides a lower bound for deterministic noise and an upper bound for random noise. In addition, We see better parameter recovery for random noise than deterministic noise. This is likely because we effectively have half as many trials for deterministic noise. In particular, while we generate two samples of random noise for each repeated game pair, we only generate one sample of deterministic noise, which by definition is the same in both of the repeated games. 
	}
	\subsubsection*{Model-based results}

	Posterior distributions over the group-level means of the deterministic and random noise standard deviation $\sigma_{det}$ and $\sigma_{ran}$ are shown in Figure \ref{fig:mb1} and Supplementary Figure S8. Consistent with our model-free results, we see that both random and deterministic noise are non-zero. \replaced[id=siyu]{Numerically,}{and that} random noise is about 2-3 times larger than the deterministic noise. \added[id=siyu]{By computing the posterior distribution of $\sigma^2_{det}/(\sigma^2_{det}+\sigma^2_{ran})$, our data suggests that 14.25\% of the variability in random exploration is accounted for by deterministic noise ([4.90\%, 28.81\%], 95\% CI).} In addition, we find that both random and deterministic noise increase with horizon. This increase was larger for random noise (mean = 7.13, 100\% of samples showed an increase in random noise with horizon) than deterministic noise (mean = 2.59, 98.64\% of samples showed an increase in deterministic noise with horizon). But intriguingly, the relative increase in both types of noise was \replaced[id=siyu]{similar}{the same} (Figure \ref{fig:ratio}). That is when we compute the relative increase in deterministic noise with horizon, $\sigma^{det}_{horizon6}/\sigma^{det}_{horizon1}$, it is \replaced[id=siyu]{very similar}{almost identical} to the relative increase in random noise with horizon $\sigma^{ran}_{horizon6}/\sigma^{ran}_{horizon1}$. 
	%BOB SAYS: CHECK NOTATION HERE AND INSERT FIGURE SHOWING RATIO

@@

	% BOB SAYS: CAN YOU TRY TO REFRAME THIS AROUND THE DETERMINISTIC STORY
	\added[id=siyu]{Randomness in exploratory behavior can be driven by both the intrinsic variability generated in the brain (show up in behavior as random noise) and by deterministic neural processes induced by the stimuli (show up in behavior as deterministic noise).} In this paper, we investigated whether random exploration is really random or whether it is driven deterministically by aspects of the stimulus we have previously ignored when measuring `decision noise'.  Using a version of the Horizon Task with repeated games, we found evidence that at least some of the noise in random exploration could be explained by such `deterministic noise.' In particular, we found that deterministic noise accounted for around 15\% of the overall variability in people's behavior.

@@

	So where does this leave randomness in random exploration? \replaced[id=siyu]{In our model, random noise is modeled as non-stimulus-driven variability, which may include truly random noise (intrinsic stochasticity in decision making) or stimulus irrelevant deterministic processes that we could not control in the lab. As a result,}{Well, } the remaining 85\% of the decision noise could be random or it could be deterministic \replaced[id=siyu]{processes that is not bound to the stimulus and that we can not control}{because we can't control everything}. In particular, while we controlled many aspects of the stimulus across repeated games (e.g. the outcomes and the order of the forced trials), we could not perfectly control {\it all} stimuli the participant received, which would vary, for example, based on exactly what they were looking at or whether they were fidgeting or scratching their nose \citep{Musall2019}. Thus conceptually, our estimate of random noise is an upper bound as these `missing' sources of deterministic noise would be interpreted as random noise in our model. Conversely, our estimate of deterministic noise is a lower bound. \added[id=siyu]{Future work is needed to identify these additional sources of deterministic noise that are not controlled in our work, for example by tracking people's pose, body movements and eye movements during the experiment.}

	In addition to the conceptual limitation in measuring deterministic noise, parameter recovery simulations suggest that our estimation method also systematically underestimates deterministic noise (see Figure \ref{fig:coverage2}, Supplementary Figure S6). From both a conceptual and methodological perspective, our model provides a lower bound of deterministic noise and an upper bound of random noise. Although there is still a considerable window for truly stochastic processes in the brain to be driving random exploration (up to 85\%), our results suggest that at least some of the apparent randomness in random exploration is not random at all.

	\added[id=siyu]{The deterministic noise hypothesis is in line with works in which neural variability can be accounted for by fluctuations in sensory inputs. For example, }MT neurons were shown to have a reproducible temporal modulation in response to a fixed random motion stimuli \citep{Bair1996}. In other words, irrelevant features in the stimuli are represented in a reliable way in the brain that could drive downstream choices in a predictable way. Of particular interest was the fact that deterministic noise in our study increased with horizon, which is a hallmark of an exploratory process and suggests that the modulation of deterministic processes may underlie random exploration.

	\added[id=siyu]{The random noise hypothesis, on the other hand, is consistent with findings of \cite{drugowitsch16}.} In particular these authors show that randomness in behavior could arise from imperfections in mental inference, which happen inside the brain, rather than in peripheral processes such as sensory processing and response selection. This suggests that a large proportion of variability in behavior may arise from computational errors in computing the correct strategy. \added[id=siyu]{Although suboptimal inference is different from simply adding random noise to a neural circuitry\citep{Pouget12}, as long as the suboptimality in neural computation is not deterministically determined by the stimuli, it is a form of random noise in our definition. }In the context of the Horizon Task, such computational errors would likely be larger in the long horizon condition as the correct course of action in these cases is much harder to compute.	
	%SIYU says the song bird literature supports that behavioral variability is reflected in neural variability. But it can't tell whether the neural variability arises from deterministic processes of external stimuli or random noise. I didn't read/remember the songbird paper, this is only based on the text below.
		%"The idea that random exploration is driven by truly stochastic processes in the brain is in line with work in the bird song literature in which variability during song learning has been tied to neural variability arising from specific areas of the brain \citep{songbird1, songbird2}"

	%this is consistent with findings of \cite{drugowitsch16}. In particular these authors show that randomness in behavior arises from imperfections in mental inference, that happen inside the brain, rather than in peripheral processes such as sensory processing and response selection. This suggests that a large proportion of noise in behavior is generated randomly and that this may arise from computational errors in computing the correct strategy. % In the context of the Horizon Task, such computational errors would likely be larger in the long horizon condition as the correct course of action in these cases is much harder to compute.

@@

	%Taken at face value, the horizon-dependent increase in random noise is consistent with the idea that random exploration is driven by intrinsic variability in the brain. 
	\added[id=siyu]{
	Regardless of whether the remaining 85\% is deterministic or random, the fact that the horizon change in the two noises are proportional to each other (Figure \ref{fig:ratio}) suggests a possible mechanism for random exploration. Specifically, a reduction in the strength with which reward drives the choice. In our decision model, choice is determined by the sign of the difference in utility $\Delta Q$ between the two options, thus, the absolute value of different terms in the model do not matter, $\Delta Q$ is only affected by the relative magnitude of reward, information, bias and noise. Mathematically, our model is equivalent to
	}
	\begin{equation*}
		\begin{split}
			\Delta Q' &= \beta \Delta Q \\
			&= \beta \Delta R + \beta A \Delta I + \beta b + \beta n_{det} + \beta n_{ran}\\
			&= \beta \Delta R + A' \Delta I + b' + n_{det}' + n_{ran}'
		\end{split}
	\end{equation*}
	\added[id=siyu]{In this equivalent form, $\beta$ determines the weight given to the reward in the decision. From a model fitting perspective, these two forms are equivalent. A decrease in $\beta$ here would be equivalent to an increase in variance of both deterministic and random noise. From a psychological perspective, however, they are quite different. An increase in both random and deterministic noise at a similar ratio relative to the reward, as observed in our experiment when horizon increases, suggests that either $\beta$ is reduced or $n_{det}'$ and $n_{ran}'$ are increased simultaneously at the same ratio. Neurally, reducing the $\beta$ corresponds to a reduction in reward sensitivity and reward coding, whereas simultaneously increasing $n_{det}'$ and $n_{ran}'$ suggests a common noise gating mechanism in which the noise filtering neural circuit is inhibited so that both random and deterministic noises are amplified. }
	%SIYU says the Ebitz paper sounds like choice signal is reduced during exploration, and p(switch|reward outcome) increases behaviorally in exploration state compared to exploitation? I am not sure it supported reward reduction?
		%Such reduced coding of reward would be consistent with \cite{ebitz17}.%In a recent report from \cite{ebitz17}, the behavioral variability of monkeys in an `explore' state was also tied to random rather than deterministic sources of noise.

@@
